---
title: "Urban Mobility - CDMX"
author: "Edgar Daniel"
page-layout: article
---


# Urban Mobility 

In the present article, we understand as Urban Mobility the join infrastructure in terms of public transportation thats is available in the city in order to understand how this blueprints interacts with the populated area of the city as well as strategic places in order to understand how people can move between places using uniquely this transportation infrastructure. 

```{r, echo = FALSE, warning=FALSE}

### Needed Libraries 
library(tidyverse)
library(knitr)
library(dplyr)
library(leaflet)
library(here)
library(sf)


# Ensure chunks use project root
opts_knit$set(root.dir = here::here())

# Paths to project root
RAW_DATA <- here("data", "raw", "urban_mobility_cdmx")
SAMPLE_DATA <- here("data", "sample")
CODE <- here("lib")


### Parameters  

# City boundaries 
CITY_SHP_DIR <- here("data", "raw", "marco_geo_2020")
SHP_DIRECTORY <- "/09_ciudaddemexico/conjunto_de_datos/"
SHP_FILE_C <- "09ent.shp"
CRS_MX <- 32614


# Stations Files 
SHP_FILE_STATION_ECOBICI = "/cicloestaciones_ecobici/cicloestaciones_ecobici.shp"
SHP_FILE_STATION_METROBUS = "/mb_shp/Metrobus_estaciones.shp"
SHP_FILE_STATION_METRO = "/stcmetro_shp/STC_Metro_estaciones_utm14n.shp"
SHP_FILE_STATION_RTP = "/rtp_shp/RTP_paradas.shp"
SHP_FILE_STATION_CABLEBUS = "/ste_shp/ste_cablebus_shp/STE_Cablebus_estaciones.shp"
SHP_FILE_STATION_TRENL = "/ste_shp/ste_tren_ligero_shp/STE_TrenLigero_estaciones_utm14n.shp"
SHP_FILE_STATION_TROLEBUS = "/ste_shp/ste_trolebus_shp/STE_Trolebus_Paradas.shp"


# Line Files 



# Rout files 



```


**Problem Statement**:

Mexico City is one of the most populous metropolitan areas in the world. As the political and economic capital of Mexico, it has been the centre of commerce and resource distribution since its foundation. Millions of people work and live here, so it is important to ask the following questions:

  - Is the transport infrastructure accessible to the whole city?
  - Is the supply of public transport sufficient to meet demand?
  - What are the expectations for a demand shock and simulation?


## Data Preparation 

The following approach was used to prepare the data: Using INEGI's 2020 Geostatistical Framework, the Mexico City street file was filtered by the relevant municipality. Based on this filter, an auto-intersection was applied to obtain the corners corresponding to the municipality's roads. 



```{r}
### Data ---------------------------------------------------------------------

## City file 
df_boundary = st_read(paste0(CITY_SHP_DIR, SHP_DIRECTORY, SHP_FILE_C))

## Read file data 
df_metro <- st_read(paste0(RAW_DATA,SHP_FILE_STATION_METRO))
df_metrobus <- st_read(paste0(RAW_DATA,SHP_FILE_STATION_METROBUS))
#df_ecobici <- st_read(paste0(RAW_DATA,SHP_FILE_STATION_ECOBICI))
#df_rtp <- st_read(paste0(RAW_DATA,SHP_FILE_STATION_RTP))
#df_cablebus <- st_read(paste0(RAW_DATA,SHP_FILE_STATION_CABLEBUS))
#df_trenl <- st_read(paste0(RAW_DATA,SHP_FILE_STATION_TRENL))
#df_trolebus <- st_read(paste0(RAW_DATA,SHP_FILE_STATION_TROLEBUS))


## Transformations 
df_metro <- st_zm(df_metro, drop = TRUE, what = "ZM")
df_metro <- st_transform(df_metro, 4326)
df_metrobus <- st_zm(df_metrobus, drop = TRUE, what = "ZM")
df_metrobus <- st_transform(df_metrobus, 4326)



```


```{r}



```




```{r}



# Final map for CDMX isochrone coverage 
map_mob <- leaflet(options = leafletOptions(minZoom = 9, zoomControl = FALSE)) |>
  addProviderTiles(providers$CartoDB.PositronNoLabels) |>

  # CDMX boundary
  addPolygons(
    data = df_boundary |> st_transform(4326),
    color = "#2b2b2b", weight = 2,
    fillOpacity = 0.1, 
    fillColor = "gray"
  ) |>
  
  # Metro Lines 
  addCircleMarkers(
      data = df_metro, 
      color = "orange", 
      radius = 4,
      stroke = FALSE,
      fillOpacity = 0.7,
      popup = ~paste0("<b>", NOMBRE, "</b><br>Linea: ", LINEA)
    ) |>
  
  # Metrobus
  addCircleMarkers(
      data = df_metrobus, 
      color = "red",
      radius = 4,
      stroke = FALSE,
      fillOpacity = 0.7,
      popup = ~paste0("<b>", NOMBRE, "</b><br>Linea: ", LINEA)
    )


map_mob 


```


