---
title: "Urban Mobility - CDMX"
author: "Edgar Daniel"
page-layout: article
---


# Urban Mobility 

In the present article, we understand as Urban Mobility the join infrastructure in terms of public transportation thats is available in the city in order to understand how this blueprints interacts with the populated area of the city as well as strategic places in order to understand how people can move between places using uniquely this transportation infrastructure. 

```{r, echo = FALSE, warning=FALSE}

### Needed Libraries 
library(tidyverse)
library(knitr)
library(dplyr)
library(leaflet)
library(here)
library(sf)
library(osrm)
library(leaflet.extras)


# Ensure chunks use project root
opts_knit$set(root.dir = here::here())

# Paths to project root
RAW_DATA <- here("data", "raw", "urban_mobility_cdmx")
SAMPLE_DATA <- here("data", "sample")
CODE <- here("lib")


source(paste0(CODE, "/distance_utils.R"))


### Parameters  

# City boundaries 
CITY_SHP_DIR <- here("data", "raw", "marco_geo_2020")
SHP_DIRECTORY <- "/09_ciudaddemexico/conjunto_de_datos/"
SHP_FILE_C <- "09m.shp"
SHP_FILE_B <- "09ent.shp"
CRS_MX <- 32614


# Stations Files 
SHP_FILE_STATION_ECOBICI = "/cicloestaciones_ecobici/cicloestaciones_ecobici.shp"
SHP_FILE_STATION_METROBUS = "/mb_shp/Metrobus_estaciones.shp"
SHP_FILE_STATION_METRO = "/stcmetro_shp/STC_Metro_estaciones_utm14n.shp"
SHP_FILE_STATION_RTP = "/rtp_shp/RTP_paradas.shp"
SHP_FILE_STATION_CABLEBUS = "/ste_shp/ste_cablebus_shp/STE_Cablebus_estaciones.shp"
SHP_FILE_STATION_TRENL = "/ste_shp/ste_tren_ligero_shp/STE_TrenLigero_estaciones_utm14n.shp"
SHP_FILE_STATION_TROLEBUS = "/ste_shp/ste_trolebus_shp/STE_Trolebus_Paradas.shp"


# Line Files 
SHP_FILE_LINES_METROBUS = "/mb_shp/Metrobus_lineas.shp"
SHP_FILE_LINES_METRO = "/stcmetro_shp/STC_Metro_lineas_utm14n.shp"


# Rout files 
METRO_DEMAND_FILE <- "/afluenciastc_desglosado_01_2026.csv"


```


**Problem Statement**:

Mexico City is one of the most populous metropolitan areas in the world. As the political and economic capital of Mexico, it has been the centre of commerce and resource distribution since its foundation. Millions of people work and live here, so it is important to ask the following questions:

  - Is the transport infrastructure accessible to the whole city?
  - Is the supply of public transport sufficient to meet demand?
  - What are the expectations for a demand shock and simulation?


## Data Preparation 

The following approach was used to prepare the data: Using INEGI's 2020 Geostatistical Framework, the Mexico City street file was filtered by the relevant municipality. Based on this filter, an auto-intersection was applied to obtain the corners corresponding to the municipality's roads. 



```{r}
### Data ---------------------------------------------------------------------

## City file 
df_boundary = st_read(paste0(CITY_SHP_DIR, SHP_DIRECTORY, SHP_FILE_B))

## Read file data 

# Stations 
df_metro_s<- st_read(paste0(RAW_DATA,SHP_FILE_STATION_METRO))
df_metrobus_s <- st_read(paste0(RAW_DATA,SHP_FILE_STATION_METROBUS))

# Lines 
df_metro_l<- st_read(paste0(RAW_DATA,SHP_FILE_LINES_METRO))
df_metrobus_l <- st_read(paste0(RAW_DATA,SHP_FILE_LINES_METROBUS))

# Isochrones 
iso_all_metro <- st_read(paste0(RAW_DATA, "/iso_10m_metro.geojson"))
iso_all_metrobus <- st_read(paste0(RAW_DATA, "/iso_10m_metrobus.geojson"))


## Transformations 
df_metro_s <- st_zm(df_metro_s, drop = TRUE, what = "ZM")
df_metro_s <- st_transform(df_metro_s, 4326)
df_metrobus_s <- st_zm(df_metrobus_s, drop = TRUE, what = "ZM")
df_metrobus_s <- st_transform(df_metrobus_s, 4326)

df_metro_l <- st_zm(df_metro_l, drop = TRUE, what = "ZM")
df_metro_l <- st_transform(df_metro_l, 4326)
df_metrobus_l <- st_zm(df_metrobus_l, drop = TRUE, what = "ZM")
df_metrobus_l <- st_transform(df_metrobus_l, 4326)


# Read MZA city data 
df_mznas <- st_read(paste0(CITY_SHP_DIR,SHP_DIRECTORY , SHP_FILE_C))



```

```{r}

# Ensure we are considering foot velocities 
options(osrm.profile = "foot")


## Metro 

# Ensure CRS is correct
df_metro <- st_transform(df_metro, 4326)

# If your server supports walking
options(osrm.profile = "foot")

iso_list <- lapply(
  seq_len(nrow(df_metro)),
  function(i) {
    osrmIsochrone(
      loc = df_metro[i, ],
      breaks = c(10)
    )
  }
)

# Combine all isochrones into one sf object
iso_all_metro <- st_union(st_make_valid(do.call(rbind, iso_list)))


```

```{r}

# Ensure we are considering foot velocities 
options(osrm.profile = "foot")


## Metro 

iso_list_mb <- vector("list", nrow(df_metrobus))

for (i in seq_len(nrow(df_metrobus))) {
  iso_list_mb[[i]] <- osrmIsochrone(loc = df_metrobus[i, ], breaks = 10)
  iso_list_mb[[i]]$CVE_EST <- df_metrobus$CVE_EST[i]
  Sys.sleep(0.2)  # small delay
}

iso_all_metrobus <- do.call(rbind, iso_list_mb)

# Combine all isochrones into one sf object
sf_use_s2(FALSE)

iso_all_metro_m <- st_transform(iso_all_metrobus, 3857) |> st_make_valid()
iso_union_m     <- st_union(iso_all_metro_m)

iso_all_metrobus <- st_transform(st_sfc(iso_union_m, crs = 3857), 4326) |> st_sf()


```


```{r}

iso_all_metro |>
  st_write(paste0(RAW_DATA, "/iso_10m_metro.geojson"),
           delete_dsn = TRUE)

iso_all_metrobus |>
  st_write(paste0(RAW_DATA, "/iso_10m_metrobus.geojson"),
           delete_dsn = TRUE)


```
Now we unify the isochrones into a same SHP object in order to make easier the intersection from covered MZNA units from the metro and metrobus stations. 

```{r}

# Merge both isochorne objects 
iso_metro_net <- st_union(iso_all_metrobus,iso_all_metro) |>
   st_transform(
     crs_iso <- st_crs(st_geometry(df_mznas))
                )


# Merge Stations 


## Get the containing MZAS 

# Centroids for MZNA 
df_mzna_cen <- df_mznas |> 
  select(CVEGEO) |> 
  mutate(
    centroid = st_centroid(geometry),
    coords = centroid |> st_coordinates(), 
    x = coords[,1], 
    y = coords[,2]  
    )


# Centroid filtering 

df_mzna_inside <- df_mzna_cen[
  lengths(st_within(df_mzna_cen$centroid, st_geometry(iso_metro_net))) > 0,
]



## Get the distance from MZNA units to nearest statins 


# Merge Metrobus and Metro Station 
df_metrored_s <-  bind_rows(
  df_metro_s |> select(SISTEMA, NOMBRE, LINEA),
  df_metrobus_s |> select(SISTEMA, NOMBRE, LINEA)
) |> 
st_transform(st_crs(st_geometry(df_mznas))) |> 
  mutate(
    centroid = st_centroid(geometry) ,
    coords = centroid |> st_coordinates(), 
    x = coords[,1], 
    y = coords[,2]  
  )






```




```{r}


# Final map for CDMX isochrone coverage 
map_mob <- leaflet(options = leafletOptions(minZoom = 9, zoomControl = FALSE)) |>
  addProviderTiles(providers$CartoDB.PositronNoLabels) |>
  
  # CDMX boundary
  addPolygons(
    data = df_boundary |> st_transform(4326),
    color = "#2b2b2b", weight = 2,
    fillOpacity = 0.1, 
    fillColor = "gray"
  ) |>
  

  # Metro Lines 
  addPolylines(
      data = df_metro_l |> st_transform(4326), 
      color = "orange", 
      opacity = 0.7,
      popup = ~paste0("<b>", RUTA, "</b><br>Linea: ", LINEA)
    ) |>
  
  # Metrobus Lines 
  addPolylines(
      data = df_metrobus_l |> st_transform(4326), 
      color = "red", 
      opacity = 0.7,
      popup = ~paste0("<b>", RUTA, "</b><br>Linea: ", LINEA)
    ) |>
  
  # Metro Stations  
  addCircleMarkers(
      data = df_metro_s |> st_transform(4326), 
      color = "orange", 
      radius = 4,
      stroke = FALSE,
      fillOpacity = 0.7,
      popup = ~paste0("<b>", NOMBRE, "</b><br>Linea: ", LINEA)
    ) |>
  
  # Metrobus
  addCircleMarkers(
      data = df_metrobus_s |> st_transform(4326), 
      color = "red",
      radius = 4,
      stroke = FALSE,
      fillOpacity = 0.7,
      popup = ~paste0("<b>", NOMBRE, "</b><br>Linea: ", LINEA)
    ) |>
  
  # Isochrones
  addPolygons(
    data = iso_all_metro |> st_transform(4326),
    color = "orange", weight = 2,
    fillOpacity = 0.25, 
    fillColor = "orange"
  ) |>

  # Isochrones
  addPolygons(
    data = iso_all_metrobus |> st_transform(4326),
    color = "salmon", weight = 2,
    fillOpacity = 0.25, 
    fillColor = "salmon"
  ) |>

    addPolygons(
    data = df_mzna_inside |> st_transform(4326),
    color = "lightgreen", weight = 0.2,
    fillOpacity = 0.1, 
    fillColor = "gray"
  )

map_mob

```






```{r}


# Final map for CDMX isochrone coverage 
leaflet(options = leafletOptions(minZoom = 9, zoomControl = FALSE)) |>
  addProviderTiles(providers$CartoDB.PositronNoLabels) |>
  
  # CDMX boundary
  addPolygons(
    data = df_boundary |> st_transform(4326),
    color = "#2b2b2b", weight = 2,
    fillOpacity = 0.1, 
    fillColor = "gray"
  ) |>

  # MZNA Layers
  addPolygons(
    data = df_mzna_inside |> st_transform(4326),
    color = "darkgreen", weight = 0.2,
    fillOpacity = 0.6, 
    fillColor = "green"
  ) 
```  
  
  
## Distance to the nearest ststion 

Other important aspect of mobility is not only the reach withing n-minute radius, but for a more comprehensive view, understand the distance from each point in the habitated city to the nearest station. 



```{r}
dist_df <- min_manhattan_to_df2(
  df1 = df_mzna_cen |> drop_na() ,
  df2 = df_metrored_s |> drop_na() ,
  id_col = "CVEGEO",
  x_col_df1 = "x",
  y_col_df1 = "y",
  x_col_df2 = "x",
  y_col_df2 = "y",
  min_buffer = 1200,
  max_buffer = 12500
)
```


```{r}

# Merge with MZNA polygonal data 
df_mzna_cen <- df_mzna_cen |> 
    left_join(dist_df |> rename(CVEGEO = id_col)
              ,by = "CVEGEO")

df_heat <- df_mzna_cen |>
  st_transform(4326) |>
  st_centroid() |>
  mutate(
    lon = st_coordinates(geometry)[, 1],
    lat = st_coordinates(geometry)[, 2]
  ) |>
  st_drop_geometry() |>
  filter(is.finite(min_dist))         

 
pal <- colorBin(
  "inferno",
  df_mzna_cen$min_dist,
  bins = 10,
  pretty = TRUE,
  na.color = "darkred",
  alpha = FALSE,
  reverse = FALSE,
  right = FALSE
)


leaflet(options = leafletOptions(minZoom = 9, zoomControl = FALSE)) |>
  setView(lng = -99.1332, lat = 19.4326, zoom = 11) |>
  addProviderTiles(providers$CartoDB.PositronNoLabels) |>

  # CDMX boundary
  addPolygons(
    data        = df_boundary |> st_transform(4326),
    color       = "#2b2b2b",
    weight      = 2,
    fillOpacity = 0.1,
    fillColor   = "gray"
  ) |>

  # MZNA layer
  addPolygons(
    data        = df_mzna_cen |> st_transform(4326),
    color       = "white",
    weight      = 0,
    fillOpacity = 1,
    fillColor   = ~pal(min_dist)
  ) |>
  addLegend(
    position = "bottomright",
    pal      = pal,
    values   = df_mzna_cen$min_dist,
    title    = "Min distance (m)",
    na.label = ">12.5Km"
  )

  
```
  
  
## Daily demand 

Lets study the daily demand of each station in order to understan mobility demand and patters inside the city. 

```{r}

df_metro_demand <- read_csv(paste0(RAW_DATA, METRO_DEMAND_FILE))
  
library(ggplot2)

df_metro_demand |>
  group_by(fecha) |>
  summarise(total = sum(afluencia)) |>
  arrange(fecha) |>
  mutate(rolling_avg = zoo::rollmean(total, k = 7, fill = NA, align = "right")) |>
  ggplot(aes(x = fecha)) +
  geom_point(aes(y = total), alpha = 0.2, size = 1, color = "gray") +
  geom_line(aes(y = rolling_avg), color = "darkblue", linewidth = 1) +
  labs(
    title    = "Metro Daily Demand",
    subtitle = "7-day rolling average",
    x        = "Date",
    y        = "Total passengers"
  ) +
  theme_minimal()
  
```
  

