---
title: "Urban Mobility - CDMX"
author: "Edgar Daniel"
page-layout: article
---


# Urban Mobility 

In the present article, we understand as Urban Mobility the join infrastructure in terms of public transportation thats is available in the city in order to understand how this blueprints interacts with the populated area of the city as well as strategic places in order to understan how people can move between places using uniquely this transportation infrastructure. 

```{r, echo = FALSE, warning=FALSE}

### Needed Libraries 
library(tidyverse)
library(knitr)
library(dplyr)
library(leaflet)

# Ensure chunks use project root
opts_knit$set(root.dir = here::here())

# Paths to project root
RAW_DATA <- here("data", "raw", "urban_mobility_cdmx")
SAMPLE_DATA <- here("data", "sample")
CODE <- here("lib")


### Parameters  

# City boundaries 
CITY_SHP_DIR <- here("data", "raw", "marco_geo_2025")
SHP_DIRECTORY <- "/09_ciudaddemexico/conjunto_de_datos/"
SHP_FILE_C <- "09e.shp"
CRS_MX <- 32614


# Stations Files 
SHP_FILE_STATION_ECOBICI = "/cicloestaciones_ecobici/cicloestaciones_ecobici.shp"
SHP_FILE_STATION_METROBUS = "/mb_shp/Metrobus_estaciones.shp"
SHP_FILE_STATION_METRO = "/stcmetro_shp/STC_Metro_estaciones_utm14n.shp"
SHP_FILE_STATION_RTP = "/rtp_shp/RTP_paradas.shp"


# Line Files 



# Rout files 



```


**Problem Statement**:

Given a Streets shapefile and a starting point, construct an algorithm that calculates the isochrone in walking or biking distance for a given time and average speed. 


## Data Preparation 

The following approach was used to prepare the data: Using INEGI's 2020 Geostatistical Framework, the Mexico City street file was filtered by the relevant municipality. Based on this filter, an auto-intersection was applied to obtain the corners corresponding to the municipality's roads. 



```{r}
### Data ---------------------------------------------------------------------

df_metro <- 

```

