---
title: "Isochrones calculator"
author: "Edgar Daniel"
page-layout: article
---


# Isochrones 

In geospatial analysis, one of the most important issues is identifying objects of interest that are within range of a given point. Think of all the cell phones that can be reached by an antenna, all the medical centers around a point in the city, and all the competitors near a store. For most of these problems, the most immediate solution is to define the range from a point in terms of distance. It is usually assumed to be Euclidean, but care must be taken because in geospatial analysis we work with projections.

```{r, echo = FALSE, warning=FALSE}
library(here)
library(knitr)
library(dplyr)

# Ensure chunks use project root
opts_knit$set(root.dir = here::here())

# Paths to project root
RAW_DATA <- here("data", "raw", "marco_geo_2020")
SAMPLE_DATA <- here("data", "sample")
CODE <- here("lib")

source(paste0(CODE, "/isochrone.R"))

# Parametros 
SHP_DIRECTORY <- "/09_ciudaddemexico/conjunto_de_datos/"
SHP_FILE_C <- "09e.shp"
CRS_MX <- 32614
```


**Problem Statement**:

Given a Streets shapefile and a starting point, construct an algorithm that calculates the isochrone in walking or biking distance for a given time and average speed. 


## Data Preparation 

The following approach was used to prepare the data: Using INEGI's 2020 Geostatistical Framework, the Mexico City street file was filtered by the relevant municipality. Based on this filter, an auto-intersection was applied to obtain the corners corresponding to the municipality's roads. 



```{r}
### Data ---------------------------------------------------------------------



```

